---
title: "Hillsborough Assessment Analysis"
author: "Ian Hodgson"
date: "2023-12-15"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(readxl)
library(knitr)

rm(list = ls())

# define tampa bay districts
tampa_bay_districts = c("HILLSBOROUGH", "PINELLAS", "PASCO")

# read in school grade data
school_grades <- read_csv('data/schoolgrades23_linked.csv') %>% 
  filter(charter_school == "NO", 
         alternative_ese_center_school == "N")
```

A statewide report, released Dec. 11, revealed that a large share of Hillsborough County's 200,000 students are performing below grade level for English and math.

Twenty-eight of the districts 219 public schools that received a grade got a D and four received an F.
That's up from seven the year before.

Six charter schools received a D or F grade, up from TK the year before.

Officials cautioned that this year's grades are not comparable to previous years and do not include adjustments for improvements over prior year's performance.

# Hillsboroughs Failing Schools

Hillsborough is among the largest districts in the state and has the highest number of low performing schools.

TK schools are inside Tampa city limits, TK are in TK...

```{r school grades, echo = F}


school_grades %>% 
  filter(district_name == "HILLSBOROUGH", 
         informational_baseline_grade_2023 %in% c('D', 'F')) %>%
  select(school_name, 
         grade_2023 = informational_baseline_grade_2023,
         ela_score = english_language_arts_achievement, 
         math_score = mathematics_achievement,
         science_score = science_achievement) %>% 
  arrange(desc(grade_2023), school_name)

```

Adjusted for student population, the district has the third-highest share of students in a D of F school among districts with more than 10,000 students.

```{r, echo = F}
school_grades %>%
  mutate(grade_d_f = (informational_baseline_grade_2023 %in% c('D', 'F')),
         students_d_f = number_of_students * grade_d_f) %>% 
  group_by(district_name) %>% 
  summarise(grade_d_f = sum(grade_d_f), 
            students_d_f = sum(students_d_f),
            students_total = sum(number_of_students)) %>% 
  mutate(share_students_d_f = round(students_d_f/students_total, 3)*100) %>% 
  # filter(students_total > 10000) %>%
  select(district_name, grade_d_f, share_students_d_f) %>% 
  arrange(desc(share_students_d_f))

## temp school output for Tom ----
school_pop <- read_xlsx("data/FLDOE/2223MembBySchoolByGrade.xlsx", 
                        sheet = 2, 
                        skip = 2,
                        na = "*") %>% 
  janitor::clean_names() %>% 
  rename(number_of_students_total = number_of_students)

school_grades %>% 
  mutate(school_number = as.numeric(school_number),
         district_number = as.numeric(district_number)) %>% 
  left_join(school_pop %>% select(district_number, school_number, number_of_students_total), 
            by = c("district_number", "school_number")) %>% 
  mutate(grade_d_f = (informational_baseline_grade_2023 %in% c('D', 'F')),
         students_d_f = number_of_students_total * grade_d_f) %>% 
  group_by(district_name) %>% 
  summarise(grade_d_f = sum(grade_d_f, na.rm = T), 
            students_d_f = sum(students_d_f, na.rm = T),
            students_total = sum(number_of_students_total, na.rm = T),
            schools_total = n()) %>% 
  mutate(share_students_d_f = round(students_d_f/students_total, 3)*100) %>% 
  # filter(students_total > 10000) %>%
  select(district_name, students_total, grade_d_f, share_students_d_f) %>% 
  arrange(desc(share_students_d_f))
```

# Schools with high shares of minority and low-income students do worse.

Statewide, schools with a higher share of minority and/or low-income students tend to perform worse.
Hillsborough has a higher share of minority students the rest of the state.
Hillsborough schools are 72% minority, on average, compared with 65% statewide.

```{r Hillsborough has a higher share of minority and low income students, echo = F}
ggplot() + 
  geom_density(data = school_grades %>% filter(district_name == "HILLSBOROUGH"),
                        aes(x = percent_of_minority_students),
                        fill = "red", 
                        alpha = .3) + 
  geom_density(data = school_grades,
                        aes(x = percent_of_minority_students),
                        fill = "blue", 
                        alpha = .3) + 
  facet_wrap(facet = "type") + 
  labs(title = "Hillsborough has a higher share minority students.") + 
  xlab("Percent of minority students")
```

Hillsborough has a similar overall share of low-income student relative to the state.
However, Hillsborough is more segregated, with a higher share being under 50% low-income or 100% low-income, relative to the rest of the state.

```{r Hillsborough has a higher share of minority and low income students, echo = F}
ggplot() + 
  geom_density(data = school_grades %>% filter(district_name == "HILLSBOROUGH"),
                        aes(x = percent_of_economically_disadvantaged_students),
                        fill = "red", 
                        alpha = .3) + 
  geom_density(data = school_grades,
                        aes(x = percent_of_economically_disadvantaged_students),
                        fill = "blue", 
                        alpha = .3) + 
  facet_wrap(facet = "type") + 
  labs(title = "Hillsborough has a higher share students from lower-income families.") + 
  xlab("Percent of economically disadvantaged students")
```

Students who are minorities and/or lower-income are more likely to attend a low-performing school in Hillsborough, relative to the rest of the state.

`r summary(school_grades %>% select(minority = percent_of_minority_students, low_income = percent_of_economically_disadvantaged_students))`

`r summary(school_grades %>% filter(district_name == "HILLSBOROUGH") %>% select(minority = percent_of_minority_students, low_income = percent_of_economically_disadvantaged_students))`

At "A" schools in Hillsborough, roughly half of students were minorities and 40% were low-income.
`r summary(school_grades %>% filter(district_name == "HILLSBOROUGH", informational_baseline_grade_2023 == "A") %>% select(minority = percent_of_minority_students, low_income = percent_of_economically_disadvantaged_students))`

At "D" and "F" schools, 91% were minority, on average.
Every low performing school in Hillsborough was 100% low-income.
`r summary(school_grades %>% filter(district_name == "HILLSBOROUGH", informational_baseline_grade_2023 %in% c("D", "F")) %>% select(minority = percent_of_minority_students, low_income = percent_of_economically_disadvantaged_students))`

```{r, echo = F}
# school_grades %>% 
#   filter(district_name == "HILLSBOROUGH",
#          informational_baseline_grade_2023 != "I") %>% 
#   mutate(low_income_num = percent_of_economically_disadvantaged_students * number_of_students / 100,
#          minority_num = percent_of_minority_students * number_of_students / 100,
#          score = case_when(
#            informational_baseline_grade_2023 %in% c("D", "F") ~ "grade D-F", 
#            informational_baseline_grade_2023  %in% c("B", "C") ~ "grade B-C", 
#            informational_baseline_grade_2023 == "A" ~ "grade A", 
#            TRUE ~ "unknown")) %>%
#   group_by(score) %>% 
#   summarise(low_income = sum(low_income_num, na.rm = T),
#             minority = sum(minority_num, na.rm = T),
#             ela_level1 = sum(ela_level1_num, na.rm = T),
#             math_level1 = sum(math_level1_num, na.rm = T),
#             total_students = sum(number_of_students)) %>% 
#   mutate(share_low_income = low_income/total_students,
#          share_minority = minority/total_students,
#          share_ela_level1 = ela_level1/total_students,
#          share_math_level1 = math_level1/total_students) %>% 
#   select(score, total_students, starts_with("share")) %>% 
#   arrange(score) %>% 
#   kable(caption = "Low performing schools have a higher share of low-income and minority students.",
#         align = "lccccc",
#         digits = 3)
```

## Hillsborough elementary schools are falling behind.

On average, Hillsborough schools perform only slightly worse than the state overall.
The greatest discrepancy was among Hillsborough elementary schools, where the average total score was 48.9% - 4.5 percentage points lower than the statewide average.

Twenty-four Hillsborough's 127 elementary schools received an D and four received an F.

```{r, echo = F}
school_grades %>% 
  filter(district_name %in% tampa_bay_districts) %>% 
  group_by(district_name, type) %>% 
  summarise(Score = mean(percent_of_total_possible_points, na.rm = T)) %>% 
  pivot_wider(names_from = "type", values_from = "Score") %>% 
  rename(District = district_name) %>% 
  select(District, Elementary, Middle, High, Combination) %>% 
  bind_rows(school_grades %>% 
              ungroup() %>% 
              group_by(type) %>% 
              summarise(Score = mean(percent_of_total_possible_points, na.rm = T)) %>% 
              mutate(district_name = "FLORIDA") %>%
              pivot_wider(names_from = "type", values_from = "Score") %>% 
              rename(District = district_name) %>% 
              select(District, Elementary, Middle, High, Combination)) %>% 
  kable(digits = 1,
        align = "lccc",
        caption = "Hillsborough elementary schools perform the worst in the Tampa Bay area, on average.")
```

However, variance in performance in higher in Hillsborough, meaning that the district's poor performing school do especially badly.
Notice how in the chart below how Hillsborough (in red) tends to skew to the left of the statewide levels (in blue), especially the district's elementary schools.

```{r, echo = F}
ggplot() + 
  geom_density(data = school_grades %>% filter(district_name == "HILLSBOROUGH"),
                        aes(x = percent_of_total_possible_points),
                        fill = "red", 
                        alpha = .3) + 
  geom_density(data = school_grades,
                        aes(x = percent_of_total_possible_points),
                        fill = "blue", 
                        alpha = .3) + 
  facet_wrap(facet = "type") + 
  labs(title = "Hillsborough has a higher share of low-performing schools.") + 
  xlab("Percent of Total Possible Points")
```

# Student Achievement

The largest component of school grades are student achievement scores.

The scores are broken up into four categories: English, math, science and social studies.
English exams are taken from grades 3-10, math grades 3-8, science grades 5 and 8 and social studies grades TK.

For each subject, students are graded on a scale from 1 to 5, with 3 and above considered performing at or above grade level.
Schools are scored from zero to 100 corresponding to the share of students that score a 3 or higher.

The scores are not adjusted for exceptionally good or poor students.

It is possible for a school to score fairly well, but have a high share of students reading at the lowest level (level 1).

```{r hillborough acievement chart, echo = F}
school_grades %>% 
  filter(district_name == "HILLSBOROUGH",
         informational_baseline_grade_2023 != "I") %>% 
  select(school_name,
         informational_baseline_grade_2023,
         english_language_arts_achievement, 
         mathematics_achievement,
         science_achievement,
         social_studies_achievement) %>% 
  rename(English = english_language_arts_achievement, 
         Math = mathematics_achievement, 
         Science = science_achievement,
         `Social Studies` = social_studies_achievement,
         Grade = informational_baseline_grade_2023) %>% 
  pivot_longer(cols = 3:6,
               names_to = "Subject", 
               values_to = "Score") %>% 
  ggplot() + 
  geom_jitter(aes(x = Score, y = Grade,
                  col = Subject), 
              alpha = .3,
              height = .15) + 
  facet_wrap(facet = "Subject", ncol = 2)
```

## English Language arts

Students from grade 3 to 10 take the English language arts assessment annually.

More than half of Hillsborough students who took the test were unable to demonstrate passing levels of proficiency and one-third performed at an "inadequate" level, indicating that they are "highly likely to need substantial support for the next grade."

Those numbers are only slightly higher than statewide averages, but what's concerning is that Hillsborough's poor performing students are more concentrated in a smaller number of failing schools.

```{r county ela levels table, echo =F}
school_grades %>% 
  filter(district_name %in% c("HILLSBOROUGH",
                              "PINELLAS", 
                              "PASCO",
                              "POLK")) %>% 
  group_by(district_name) %>% 
  summarise_at(.vars = vars(number_of_students, 
                            ela_level1_num, 
                            ela_level2_num, 
                            ela_level3_num, 
                            ela_level4_num, 
                            ela_level5_num),
               .funs = function(x){sum(x, na.rm = T)}) %>% 
  mutate(ela_level1 = ela_level1_num/number_of_students*100,
         ela_level2 = ela_level2_num/number_of_students*100,
         ela_level3 = ela_level3_num/number_of_students*100,
         ela_level4 = ela_level4_num/number_of_students*100,
         ela_level5 = ela_level5_num/number_of_students*100) %>% 
  select(District = district_name, 
         `Level 1` = ela_level1,
         `Level 2` = ela_level2,
         `Level 3` = ela_level3,
         `Level 4` = ela_level4,
         `Level 5` = ela_level5) %>% 
  kable(digits = 1,
        caption = "English languge arts scores by level")
```

```{r county ela levels table, echo =F}
school_grades %>% 
  filter(district_name %in% c("HILLSBOROUGH")) %>% 
  group_by(informational_baseline_grade_2023) %>% 
  summarise_at(.vars = vars(number_of_students, 
                            ela_level1_num, 
                            ela_level2_num, 
                            ela_level3_num, 
                            ela_level4_num, 
                            ela_level5_num),
               .funs = function(x){sum(x, na.rm = T)}) %>% 
  mutate(ela_level1 = ela_level1_num/number_of_students*100,
         ela_level2 = ela_level2_num/number_of_students*100,
         ela_level3 = ela_level3_num/number_of_students*100,
         ela_level4 = ela_level4_num/number_of_students*100,
         ela_level5 = ela_level5_num/number_of_students*100) %>% 
  select(Grade = informational_baseline_grade_2023, 
         `Level 1` = ela_level1,
         `Level 2` = ela_level2,
         `Level 3` = ela_level3,
         `Level 4` = ela_level4,
         `Level 5` = ela_level5) %>% 
  kable(digits = 1,
        caption = "Hillsborough English languge arts scores by school grade")
```

The issue in Hillsborough is that the poor performing students are concentrated among low-rated schools.

```{r, echo = F}
ggplot() + 
  geom_density(data = school_grades %>% filter(district_name == "HILLSBOROUGH"),
                        aes(x = english_language_arts_achievement),
                        fill = "red", 
                        alpha = .3) + 
  geom_density(data = school_grades,
                        aes(x = english_language_arts_achievement),
                        fill = "blue", 
                        alpha = .3) + 
  facet_wrap(facet = "type") + 
  labs(title = "Hillsborough has a higher share of low performing elementary, middle\nand combination schools, relative to the state overall.",
       subtitle = "English Language Arts Achievement")

```

There were 138 schools in Hillsborough where more than half of students were reading below grade level.
There were 32 where more than half of students were reading at level 1.

```{r, echo = F}
ggplot() + 
  geom_density(data = school_grades %>% filter(district_name == "HILLSBOROUGH"),
                        aes(x = ela_level1),
                        fill = "red", 
                        alpha = .3) + 
  geom_density(data = school_grades,
                        aes(x = ela_level1),
                        fill = "blue", 
                        alpha = .3) + 
  facet_wrap(facet = "type") + 
  labs(title = "Hillsborough has a higher share of students reading at the lowest level.",
       subtitle = "Share of students reading at level 1")
```

At two of the worst performing elementary schools in the state - Potter Elementary and Just Elementary - three in five students were reading at the lowest level.

Even well performing schools had high shares of very poor readers.
Three middle schools - Greco, Pierce and Buchanan Middle - all had more than 60% of students reading at the lowest level despite earning a "C" grade overall.
Fourteen schools earned a "C" grade despite having more than half of students reading at level 1.

```{r, echo = F}
school_grades %>% 
  filter(district_name == "HILLSBOROUGH",
         informational_baseline_grade_2023 != "I") %>% 
  rename(Grade = informational_baseline_grade_2023) %>% 
  ggplot() + 
  geom_jitter(aes(x = ela_level1, 
                  y = Grade, 
                  col = Grade), 
              height = .15,
              width = 0,
              alpha = .5) + 
    facet_wrap(facet = "type") + 
  geom_vline(xintercept = .51,
             alpha = .5) + 
  labs(title = "More than half of students are level 1 readers at 32 Hillsborough schools",
       subtitle = "")
```

## Mathematics

Students from grades 3 to 8 take the mathematics exam each year.

```{r county math levels table, echo =F}
school_grades %>% 
  filter(district_name %in% c("HILLSBOROUGH",
                              "PINELLAS", 
                              "PASCO",
                              "POLK")) %>% 
  group_by(district_name) %>% 
  summarise_at(.vars = vars(number_of_students_math, 
                            math_level1_num, 
                            math_level2_num, 
                            math_level3_num, 
                            math_level4_num, 
                            math_level5_num),
               .funs = function(x){sum(x, na.rm = T)}) %>% 
  mutate(math_level1 = math_level1_num/number_of_students_math*100,
         math_level2 = math_level2_num/number_of_students_math*100,
         math_level3 = math_level3_num/number_of_students_math*100,
         math_level4 = math_level4_num/number_of_students_math*100,
         math_level5 = math_level5_num/number_of_students_math*100) %>% 
  select(District = district_name, 
         `Level 1` = math_level1,
         `Level 2` = math_level2,
         `Level 3` = math_level3,
         `Level 4` = math_level4,
         `Level 5` = math_level5) %>% 
  kable(digits = 1,
        caption = "Mathematics scores by level")
```

```{r county math levels table, echo =F}
school_grades %>% 
  filter(district_name == "HILLSBOROUGH") %>% 
  group_by(informational_baseline_grade_2023) %>% 
  summarise_at(.vars = vars(number_of_students_math, 
                            math_level1_num, 
                            math_level2_num, 
                            math_level3_num, 
                            math_level4_num, 
                            math_level5_num),
               .funs = function(x){sum(x, na.rm = T)}) %>% 
  mutate(math_level1 = math_level1_num/number_of_students_math*100,
         math_level2 = math_level2_num/number_of_students_math*100,
         math_level3 = math_level3_num/number_of_students_math*100,
         math_level4 = math_level4_num/number_of_students_math*100,
         math_level5 = math_level5_num/number_of_students_math*100) %>% 
  select(Grade = informational_baseline_grade_2023, 
         `Level 1` = math_level1,
         `Level 2` = math_level2,
         `Level 3` = math_level3,
         `Level 4` = math_level4,
         `Level 5` = math_level5) %>% 
  kable(digits = 1,
        caption = "Hillsborough Mathematics scores by school grade")
```

**Questions**

-   Weighted average or not?

-   Group D and F schools?

-   
