---
title: "Hillsborough School Grades Checking"
author: "Ian Hodgson"
date: "2024-03-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, 
                      warning = F, 
                      message = F,
                      root.dir = "/Users/ihodgson/Documents/GitHub/hillsborough_grades/")
knitr::opts_knit$set(echo = F, 
                      warning = F, 
                      message = F,
                      root.dir = "/Users/ihodgson/Documents/GitHub/hillsborough_grades/")
library(tidyverse)
library(janitor)
library(readxl)
library(knitr)
library(gt)
```

```{r import and build necessary datasets, include = F}
# School grades ----
school_grades <- read_csv('data/schoolgrades23_linked.csv') %>% 
  # NOTE Monroe Middle was changed late from an "I" for incomplete to a "D" school.
  mutate(informational_baseline_grade_2023 = ifelse(school_number == "2362", "D",
                                                    informational_baseline_grade_2023)) %>% 
  filter(charter_school == "NO", 
         alternative_ese_center_school == "N") %>% 
  ## For most of these stats we want to consider D/F schools together and K-8 schools grouped with middle schools.
  mutate(Grade = ifelse(informational_baseline_grade_2023 %in% c("D", "F"), 
                        "D/F",
                        informational_baseline_grade_2023),
         type_orig = type, 
         type = ifelse(type == "Combination", "Middle", type)) %>% 
  select(-number_of_students)

df_hillsborough <- school_grades %>% 
  filter(district_name == "HILLSBOROUGH",
         Grade == "D/F")

## Student populations ----
enrollment_2223 <-read_xlsx("data/FLDOE/Enrollment/2223MembBySchoolByGrade.xlsx",
                             sheet = 2,
                             skip = 2) %>% 
  janitor::clean_names() %>% 
  # filter(district == "HILLSBOROUGH") %>% 
  mutate(sch_number = as.numeric(school_number),
         number_of_students = as.numeric(number_of_students)) %>% 
  select(district, sch_number, total_enrollment = number_of_students)

grade_enrollment <- school_grades %>% 
  mutate(school_number = as.numeric(school_number)) %>% 
  left_join(enrollment_2223, 
            by = c("district_name" = "district",
                   "school_number" = "sch_number")) %>% 
  mutate(df_students = (Grade == "D/F")*total_enrollment) %>% 
  group_by(district_name) %>% 
  summarise(schools = n(), 
            df_schools = sum(Grade == "D/F", na.rm = T),
            total_students = sum(total_enrollment, na.rm = T),
            df_students = sum(df_students, na.rm = T)) %>% 
  arrange(desc(total_students)) %>% 
  head(n = 10)

```

#### Each day, 18,000 students take their seats inside Hillsborough County’s most struggling schools, enough to fill a small city.

``` {r 18,000 kids}
enrollment_2223 %>% 
  filter(district == "HILLSBOROUGH",
         sch_number %in% as.numeric(df_hillsborough$school_number)) %>% 
  summarise(number_of_students = sum(total_enrollment, na.rm = T))
```


#### 33 Hillsborough elementary and middle schools received a D or F grade from the state last year. That’s more than double the year before, and far more than any other county in Florida. Palm Beach County came closest with 23 D and F schools. Florida’s largest district, Miami-Dade, had five.


```{r school grades, include = T}
school_grades %>% 
  filter(district_name %in% c("MIAMI-DADE",
                              "BROWARD", 
                              "ORANGE",
                              "HILLSBOROUGH",
                              "PALM BEACH",
                              "DUVAL",
                              "POLK",
                              "LEE",
                              "PINELLAS", 
                              "PASCO")) %>%
  group_by(district_name) %>% 
  summarise(df_2023 = sum(informational_baseline_grade_2023 %in% c("D", "F")),
            df_2022 = sum(grade_2022 %in% c("D", "F"))
            # ,
            # df_2019 = sum(grade_2019 %in% c("D", "F")), 
            # df_2018 = sum(grade_2018 %in% c("D", "F")),
            # df_2017 = sum(grade_2017 %in% c("D", "F")),
            # df_2016 = sum(grade_2016 %in% c("D", "F")),
            # df_2015 = sum(informational_baseline_grade_2015 %in% c("D", "F"))
            ) %>% 
  gt() %>% 
  tab_header("Hillsborough had 33 D/F-rated schools in 2022-23, twice the number as the previous year.")

```
#### But that’s not the case at the 33 D and F schools. A third of their teachers are inexperienced on average, nearly triple he percentage for A-rate schools.

**Source: Florida Department of Education school report cards**

``` {r inexperienced teachers}
inexperienced <- read_xlsx("data/FLDOE/School Reportcard - Inexperienced Teachers.xlsx", 
                           sheet = "A Schools") %>%
  clean_names() %>% 
  mutate(Grade = "A")
  bind_rows(read_xlsx("data/FLDOE/School Reportcard - Inexperienced Teachers.xlsx", 
                           sheet = "DF Schools") %>% 
              clean_names() %>% 
              mutate(Grade = "D/F"))

inexperienced %>% 
  filter(type != "High") %>% 
  group_by()

```
